Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-10-28 paper/paper.tex
editted paper and some of the labelling on plots after revising Davids comments and reading through it myself Tested everything on crystal.

\section{Introduction}
%% \fixme{Huang and those who reference him use the terms 'numerical
%%   deterministic reaction-diffusion model' and 'stochastic version of
%%   this model' for the stochastic.}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in prohibiting
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which ineracts with another protein, MinE, that
is neccesary for MinD oscillation.~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid,
  meacci2005min, raskin1999minde}.

Previous studies have shown experimentally that the MinD protein
system is capable of exhibiting oscillations in round
shapes~\cite{fange2006noise} as well as in connected three pronged
tube shapes~\cite{varma2008min}, in which the oscillations seem to
seek out the extreme poles in the
cell~\cite{juarez2010changes,corbin2002exploring}.  One conclusion of
these studies has been that the MinD system robustly forms bipolar
oscillations regardless of variations in cellular shape.

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2010bacteria, mannik2009bacterial}. They have
experimentally forced E. Coli cells into silicon microfabricated
channels of $0.25\micron$ thickness. Upon entering the channels, the
cells undergo a mechanical deformation in which they both widen and
lengthen within the plane of the channel.  This deformation results in
very wide (they can reach widths of over $5\micron$) flattened cells
that when viewed from the top down have irregular and asymmetric
shapes.  While these cells are still able to divide into surprisingly
equal volumes, the MinD oscillations in these cells are irregular,
both temporally and spatially. Seen with flourescent microscopy (after
using common genetic techniques to allow the cells to flouresce), the
MinD seem to maximize in multiple locations within the cell, in a
seemingly random sequence. This allows for an opportunity to test
previously studied MinD simulation models against more extreme
experimental cases than have been seen thus far.

A number of models of the MinD protein system have been developed,
which accurately describe the dynamics using simulations of protein
reaction and diffusion.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachement of minD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.  A number of studies have used an approach similar in
that they do not rely on the ability of MinD to move along the walls
and cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically treat molecular interaction
and diffusion~\cite{fange2006noise, kerr2006division}.  These studies
largely confirm the results of Huang's model, when applied to the
wild-type phenotype.  However, the stochastic models more successfully
predict experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and enable prediction
of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E
\end{multline}
\begin{multline}
  \frac{\partial \sigma_D}{\partial t} = -k_E\sigma_D\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE}
\end{multline}
where $\rho$ is cytoplasmic protein density ($\micron^{-3}$), $\sigma$
is membrane bound density ($\micron^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ th diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ the rate of cytoplasmic MinE binding to
membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall.  The Dirac delta function
$\delta(d_w)$ has units of $\micron^{-1}$ and is zero everywhere except
at the wall.  Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} are
only relevent at the membrane because the membrane-bound density
values have no meaning in the cytoplasm.

@@

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$, respectively. We
construct a 3D grid in cartesian coordinates, with a grid spacing of
.05\micron.

We have performed both a basic finite analysis, numerical,
deterministic model simulation that is spatially and temporally
discrete, and a stochastic analysis that is spatially discrete but
continuous in time.  Our stochastic model follows the work of
Kraus~\cite{kraus1996crosstalk} which in turn follows a method
introduced by Gillespie~\cite{gillespie1977exact}.

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in
several cell shapes and sizes.  Here we present a selection of these,
beginning with naturally occuring pill-shaped cells, followed by a
number of flattened out shapes which reflect the experiments of Mannik
\emph{et al.}, in which bacteria are confined within a thin slit of
height $0.25\micron$ ~\cite{mannik2012robustness}.  Viewed from the
top down the cells have the shapes described below and viewed from the
side they have at their edges a semicircular protrusion (one may
imagine the shape of a pancake).
%
In this paper we focus on four specific flattened cell shapes.  Two of
these shapes replicate those published by Mannik, and the other two
are `stadium' shapes that respectively have the same aspect ratio,
thickness, and volume as the two Mannik shapes.  Viewed from the top
down, these stadium shapes appear as rectangles with semi-circular end
caps on the long axis ends.

@@

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-2.5em}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-3em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of a $2\micron$ wavelength.  Cells with dimensions longer
than this show spontaneous oscillatory behavoir in this
direction, while cells that are shorter relax into a motionless steady
state.  We have similarly performed a stability analysis in an
infinite slab with a thickness equal to that of our flattened cells
($0.25\micron$) and have found that an equivalent stability limit for
half wavelength is $2.13\micron$. As expected, when decreasing the
lengths and widths of our simulated flattened cells so that the
longest distance across the cell is less than this length, the cells
stop exhibiting any oscillatory behavior.  The deterministic model
relaxes into a motionless state and the stochastic model exhibits
random fluctuations without spatial oscillations.

@@

Figure~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. The cells shown are
$4\micron$ in length, measured from end to end.  We have `smeared out'
the stochastic concentrations in a manner meant to reproduce the
images shown by diffraction limited flourescence microscopy.  We do so
using the two dimensional gaussian approximation developed by Zhang
\emph{et al.}~\cite{zhang2007gaussian}.  In this approximation we use
a numerical aperture value of 1.3, which is the same as used by Mannik
experimentally, and a wavelength of $650nm$.  Each frame is 2.5
seconds ahead of the last, and each image shows the concentration of a
given state of protein (of the five described in the reaction model)
summed over the coordinate orthogonal to the page.

@@

During the formation of these rings, MinE that is diffusing in the
cytoplasm in the upper portion of the cell will naturally diffuse
downward, where there is membrane-bound MinD to react with, leading to a
depletion of MinE in the upper portion of the cell.  As the MinE ring
converges upon the lower end of the cell, MinD that has been released
is able to diffuse upward, past the ring, while still in its MinD:ADP
state.  After it undergoes nucleotide exchange, resulting in MinD:ATP,
it is ready to accumulate on the walls in at the top of the
cell, where the MinE has been depleted.  This can be seen in seconds
10 through 20 in both models, followed by the subsequent MinE ring
formation and movement upward (beginning the same process in the
opposite direction) can be seen in seconds 15 through 23.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill shaped cell, shown against
    the correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and decay time of 307 seconds,
    so that it takes roughly 7.8 periods for the behavoir to
    decohere.}
  \label{corr-pill}
\end{figure}

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is less spatially regular than
the deterministic model predicts.  In fact, when viewed as a movie
(see supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom sections of a cell that has been
divided into thirds along its long axis.  These plots help us in
studying the periods and the regularity and stability of oscillations.
These curves are well fit with a simple exponential decoherence model,
given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the determinisitic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

As describe in Section~\ref{sec:model-method-shapes}, we will focus on
simulation of four illustrative flattened cell shapes: two shapes
created to replicate those published by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
cellular shape irregularity and assymetry, and that of overall size
and aspect ratio.  We note that shape \emph{B} is more irregular than
shape \emph{A}, and in particular features a prominant region of
concavity on its left-hand side.  For each of these shapes (in
addition to the wild-type pill shape discussed previously) we have
simulated in excess of 15000 seconds of evolution of the MinD
system using the deterministic version of Huang's
model~\cite{huang2003dynamic}, as well as the stochastic model.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the mannik \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  For mannik \emph{A}, the stochastic
    model shows an oscillation period of 53.1 seconds and decay time
    of 248 seconds, so that it takes roughly 4.7 periods for the
    behavoir to decohere.  For the stadium \emph{A}, the same model
    shows an oscillation period of 51 seconds and decay time of 313
    seconds, so that it takes roughly 6.1 periods for the behavoir to
    decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

The predictions of the stochastic model are roughly similar to the
experimental results in terms of irregularity of the locations of
maxima, although there are locations on the edges of the cells where
experiment shows maxima occuring that we never see in our simulations,
suggesting that the model does not precisely reflect the experiment.
We also note that the stadium shapes in Fig.~\ref{randst-plot-ave}
appear to have maxima location irregularity that is qualitatively
similar to our mannik shapes and to experiment.

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
irregularity versus overall thickness and aspect ratio in developing
an understanding of the origin of the experimentally observed
behavior.  Although in the movies the stadiums appear to display, on
average over long time scales, a more regular bipolar oscillation than
the irregular mannik shapes, the limited time range of the arrow plots
(350 seconds) is inadequate when studying this difference in behavoir.
We therefore return to the correlation function between the number of
proteins in a segment at the top of the cell and the number of
proteins in a segment at the bottom of the cell.  For each shape we
computed these correlation functions using the entirety of our over
15000 seconds of simulation data. Fig.~\ref{fig:corr-pancake-A} shows
this plot for mannik \emph{A} and stadium \emph{A}.  E. Coli cell
doubling times vary widely under alternate conditions, but will often
be between 20 minutes to 100 minutes~\cite{pierucci1972chromosome},
while flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times that range from 0 up to 600, a biologically relevent
timescale.

>>>>

\section{Introduction}
%% \fixme{Huang and those who reference him use the terms 'numerical
%%   deterministic reaction-diffusion model' and 'stochastic version of
%%   this model' for the stochastic.}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in such a fashion as to exhibit pole to pole oscillatory
behavoir.~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{fange2006noise} as well as in connected three pronged
tube shapes~\cite{varma2008min}, in which the oscillations seem to
seek out the extreme poles in the
cell~\cite{juarez2010changes,corbin2002exploring}.  One conclusion of
these studies has been that the MinD system robustly forms bipolar
oscillations regardless of variations in cellular shape.

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2010bacteria, mannik2009bacterial}. They have
experimentally forced \emph{E. coli} cells into microfabricated
silicon channels of $0.25\micron$ thickness. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are irregular, both temporally and spatially. Seen with
flourescent microscopy, the MinD maximize in multiple locations within
the cell, in a seemingly random sequence. These experiments allow for
an opportunity to test MinD simulation models against more extreme
cases than have been seen thus far.

A number of models of the MinD protein system have been developed,
which accurately describe the dynamics using simulations of
reaction-diffusion equations.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachement of minD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.  A number of studies have used an approach similar in
that they do not rely on the ability of MinD to move along the walls
and cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically treat molecular interaction
and diffusion~\cite{fange2006noise, kerr2006division}\fixme{talk about
  this awkward sentence}.  These studies largely confirm the results
of Huang's model when applied to the wild-type, pill-shaped phenotype.
However, the stochastic models more successfully predict
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and enable prediction
of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E
\end{multline}
\begin{multline}
  \frac{\partial \sigma_D}{\partial t} = -k_E\sigma_D\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE}
\end{multline}
where $\rho$ is cytoplasmic protein density ($\micron^{-3}$), $\sigma$
is membrane bound density ($\micron^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ are the diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ is the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ is the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ is the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ is the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ is the rate of cytoplasmic MinE binding
to membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall.  The Dirac delta function
$\delta(d_w)$, which we need to describe the location of the membrane,
has units of $\micron^{-1}$ and is zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} are only relevent
at the membrane because the membrane-bound density values have no
meaning in the cytoplasm.

@@

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$, respectively. We perform
simulations with a 3D grid in cartesian coordinates that has a grid
spacing of .05\micron.

We have performed both a numerical, deterministic model simulation
that is spatially and temporally discrete, and a stochastic simulation
that is spatially discrete but continuous in time.  Our stochastic
model follows the work of Kraus~\cite{kraus1996crosstalk} which in
turn follows a method introduced by
Gillespie~\cite{gillespie1977exact}.

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we have modeled the Min system
in several cell shapes and sizes.  Here we present a selection of
these, beginning with naturally occuring pill-shaped cells, followed
by a number of flattened out shapes which reflect the experiments of
Mannik \emph{et al.}, in which bacteria are confined within a thin
slit of height $0.25\micron$ ~\cite{mannik2012robustness}.  Viewed
from the top down the cells have the shapes described below and viewed
from the side they have at their edges a semicircular protrusion (one
may imagine the shape of a pancake).
%
In this paper we focus on four specific flattened cell shapes.  Two of
these shapes replicate those published by Mannik, and the other two
are `stadium' shapes that respectively have the same aspect ratio,
thickness, and volume as the two Mannik shapes.  Viewed from the top
down, these stadium shapes appear as rectangles with semi-circular end
caps on the long axis ends.

@@

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-1.5em}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-3em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of a $2\micron$ half wavelength.  Cells with dimensions
longer than this show spontaneous oscillatory behavoir in this
direction, while cells that are shorter relax into a motionless steady
state.  We have similarly performed a stability analysis in an
infinite slab with a thickness equal to that of our flattened cells
($0.25\micron$) and have found that an equivalent stability limit for
half wavelength is $2.13\micron$. As expected, when decreasing the
lengths and widths of our simulated flattened cells so that the
longest distance across the cell is less than this length, the cells
stop exhibiting any oscillatory behavior.  The deterministic model
relaxes into a motionless state and the stochastic model exhibits
random fluctuations without spatial oscillations.

@@

Figure~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. Deterministic simulation
data is shown above and stochastic simulation data is shown below.
The cells shown are $4\micron$ in length, measured from end to end.
We have `smeared out' the stochastic concentrations in a manner meant
to reproduce the images shown by diffraction limited flourescence
microscopy.  We do so using the two dimensional gaussian approximation
developed by Zhang \emph{et al.}~\cite{zhang2007gaussian}.  In this
approximation we use a numerical aperture value of 1.3, which is the
same as used by Mannik experimentally, and a wavelength of $650nm$.
Each frame is 2.5 seconds ahead of the last, and each image shows the
concentration of a given state of protein (of the five described in
the reaction model) summed over the coordinate normal to the page.

@@

During the formation of these rings, cytoplasmic MinE is diffusing in
the upper portion of the cell and will naturally progress downward,
where there is membrane-bound MinD to react with, leading to a
depletion of MinE in the upper portion of the cell.  As the MinE ring
converges upon the lower end of the cell, MinD that has been released
is able to diffuse upward, past the ring, while still in its MinD:ADP
state and unable to bind to the membrane.  After it undergoes
nucleotide exchange, resulting in MinD:ATP, it is ready to accumulate
on the walls in at the top of the cell, where the MinE has been
depleted.  This can be seen in seconds 10 through 20 in both models,
followed by the subsequent MinE ring formation and movement upward
(beginning the same process in the opposite direction) that can be
seen in seconds 15 through 23.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and decay time of 307 seconds.}
  \label{corr-pill}
\end{figure}

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is less spatially regular than
the deterministic model predicts.  In fact, when viewed as a movie
(see supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the determinisitic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

As described in Section~\ref{sec:model-method-shapes}, we will focus
on simulation of four illustrative flattened cell shapes: two shapes
created to replicate those shapes observed experimentally by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
cellular shape irregularity and assymetry, and that of overall size
and aspect ratio.  We note that shape \emph{B} is more irregular than
shape \emph{A}, and in particular features a prominant region of
concavity on its left-hand side.  For each of these shapes (in
addition to the wild-type pill shape discussed previously) we have
simulated in excess of 15000 seconds (over four hours) of evolution of
the MinD system using the deterministic version of Huang's
model~\cite{huang2003dynamic}, and much more than this using the
stochastic model. \fixme{is this sentence ok?}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  For shape \emph{A}, the stochastic
    model shows an oscillation period of 53.1 seconds and decay time
    of 246 seconds, so that it takes roughly 4.6 periods for the
    behavoir to decohere.  For the stadium \emph{A}, the same model
    shows an oscillation period of 51 seconds and decay time of 323
    seconds, so that it takes roughly 6.3 periods for the behavoir to
    decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

The predictions of the stochastic model are roughly similar to the
experimental results in terms of irregularity of the locations of
maxima, although there are locations on the edges of the cells where
experiment shows maxima occuring that we never see in our simulations,
suggesting that the model does not precisely reflect the experiment.
We also note that the stadium shapes in Fig.~\ref{randst-plot-ave}
appear to have maxima location irregularity that is qualitatively
similar to our Mannik shapes and to experiment.

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
cell shape irregularity versus overall size and aspect ratio in
developing an understanding of the origin of the experimentally
observed behavior.  Although in the movies the stadiums appear to
display, on average over long time scales, a more regular bipolar
oscillation than the irregular mannik shapes, the limited time range
of the arrow plots (350 seconds) is inadequate when studying this
difference in behavoir.  We therefore return to the correlation
function between the number of proteins in a segment at the top of the
cell and the number of proteins in a segment at the bottom of the
cell.  For each shape we computed these correlation functions using
the entirety of our over 15000 seconds of simulation
data. Fig.~\ref{fig:corr-pancake-A} shows this plot for shape \emph{A}
and stadium \emph{A}.  E. Coli cell doubling times vary widely
depending on environmental conditions, but will often be between 20
minutes to 100 minutes~\cite{pierucci1972chromosome}, while
flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times up to 600 seconds, a biologically relevent
timescale.

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  At
$\tau=0$, the stochastic correlation is already seen to be lower than
the deterministic, reflecting the fact that the stochastic data
sometimes shows maxima or minima at both ends simultaneously.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.  This moderate difference in coherence
time shows that the regularity in cell oscillation is not overly
sensitive to the regularity and symmetry of the shape.  It seems that
it is the size and aspect ratio that are more important in dictating
the regularity of bipolar oscillation.
